---
title: "Week 4 Exercises"
author: "J.Blondeau"
date: "2023-09-09"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}

```

This week we will continue wrangling data in the Tidyverse.  The following are some very common functions you will use in your everyday data world.

- merge()
- mutate()
- case_when()
- group_by()
- summarize()


All the exercises for this week use the "PSU_Fish_Density.csv" and the "taxonomic_data.csv" data sets.

1. Add the common name (only common name) of each species in the density data set as a new field.  
    - Reorder the columns to be more readable.
    
```{r}
density<-read.csv("data/PSU_Fish_Density.csv")
tax<-read.csv("data/taxonomic_data.csv")

library(tidyverse)

density1<-left_join(density,tax, by="SPECIES_CD")

density1<-density1%>%select(YEAR, REGION, STRAT, PROT, PRIMARY_SAMPLE_UNIT,SPECIES_CD,COMNAME,density)
```


2. Add a new field to the density data set named "fishery_target" at set all to TRUE
```{r}
density1<-density1%>%
  mutate(fishery_target=TRUE)
```

3. Add a new field named "group" at set the value based on fish species where...
    - grouper = Coney, Red hind, 
    - sanpper = Mutton, Gray, Yellowtail
    - parrotfish = Stoplight
    - other = Triggerfish, Hogfish
```{r}
density1<-density1%>%
  mutate(group=case_when(COMNAME%in% c("coney","red hind")~"grouper",
                  COMNAME%in%c("mutton snapper","gray snapper","yellowtail snapper")~"snapper",
                      COMNAME%in%c("stoplight parrotfish")~"parrotfish",
                      COMNAME%in%c("triggerfish","hogfish")~"other"
                        ))

```
    
4. Using the group_by function, how many unique PRIMARY_SAMPLE_UNITS were sampled in each YEAR?
```{r}

psu<-density1%>%group_by(YEAR)%>%
  summarise(length(unique(PRIMARY_SAMPLE_UNIT)))
#237 in 2017, 322 in 2019, 165 in 2021

```

5. How many unique PRIMARY_SAMPLE_UNITS were sampled in each YEAR and PROT combination?
```{r}
psuprot<-density1%>%group_by(YEAR,PROT)%>%
  summarise(length(unique(PRIMARY_SAMPLE_UNIT)))
```

6. How many unique PRIMARY_SAMPLE_UNITS were sampled in each YEAR, PROT and STRAT combination?
```{r}
psu5<-density1%>%group_by(YEAR,PROT,STRAT)%>%
  summarise(length(unique(PRIMARY_SAMPLE_UNIT)))
```

7. What is the difference between the following two expressions?  The summarize and mutate calls after the group_by do very different things.  In what situations would you use them?
    
    - data %>% # data here is the "PSU_Fish_Density" data set
        <br></br>
        group_by(YEAR) %>% 
        <br></br>
        summarize( n = n_distinct(PRIMARY_SAMPLE_UNIT))

    - data %>%
        <br></br>
        group_by(YEAR) %>% 
        <br></br>
        mutate( n = n_distinct(PRIMARY_SAMPLE_UNIT))
        
        
8. Create a new dataframe that shows mean density of each species per year...hint
    - YEAR  | SPECIES_CD  | meanDensity
      ----- | ----------  | -----------
      2017  | BAL VETU    |    0.224   
      2019  | CEP FULV    |    0.685
```{r}
mean<-density1%>%group_by(YEAR,SPECIES_CD)%>%
  summarise(mean(density))
```
      
9. Create a new dataframe that shows mean density of each species in each PROT per year...hint
    - YEAR  | SPECIES_CD  | PROT | meanDensity
      ----- | ----------  | -----|-----------
      2017  | BAL VETU    | 0    |0.229  
      2019  | CEP FULV    | 1    |0.308
```{r}
mean1<-density1%>%group_by(YEAR, SPECIES_CD,PROT)%>%
  summarise(mean(density))
```

10. In the "PSU_Fish_Density.csv" data set, the PROT field refers to sites that are inside the VI National Park (PROT = 1) and sites that are outside the Park (PROT = 0).  How many of the 8 species had higher densities inside the National Park?

```{r}
mean3<-density1%>%group_by(PROT, SPECIES_CD)%>%
  summarise(mean(density))

```

